Impact of Training Variability on Visuomotor Function Learning and Extrapolation

### 2024-07-09 Manuscript/manuscript.qmd
final edits from Rob's doc

---
title: Impact of Training Variability on Visuomotor Function Learning and Extrapolation
author:
- name: Thomas E. Gorman
  affiliations: 
  - name:  Department of Psychological and Brain Sciences, Indiana University-Bloomington, USA
    affiliation-url: https://psych.indiana.edu/
  - name:  Program in Cognitive Science, Indiana University-Bloomington, USA
    affiliation-url: https://cogs.indiana.edu/index.html
  url: https://tegorman13.github.io/
  email: tegorman@iu.edu
  orcid: "0000-0001-5366-5442"
- name: Robert L. Goldstone
  affiliation: Department of Psychological and Brain Sciences, Indiana University-Bloomington
  affiliation-url: https://psych.indiana.edu/
  url: https://pc.cogs.indiana.edu/
  email: rgoldsto@indiana.edu
  orcid: "0000-0001-8357-8358"
# abstract: |
#   In project 1, we applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, we will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. Project 2 also places a greater emphasis on extrapolation performance following training - as varied training has often been purported to be particularly beneficial in such situations. 
# keywords:
#   - Learning Generalization
#   - Function Learning
#   - Visuomotor learning
#   - Training Variability
# page-layout: full
# code-repo: "Access the code, data, and analysis at <https://github.com/tegorman13/htw>"
toc: false
---

@@

Our findings also diverge from the two previous studies to cleanly manipulate the variability of training items in a function learning task [@deloshExtrapolationSineQua1997; @vandamMappingShapeVisuomotor2015], although the varied training condition of @vandamMappingShapeVisuomotor2015 also exhibited less learning, neither of these previous studies observed any difference between training conditions in extrapolation to novel items. Like @deloshExtrapolationSineQua1997 , our participants exhibited above chance extrapolation/discrimination of novel items, however they observed no difference between any of their three training conditions. A noteworthy difference difference between our studies is that @deloshExtrapolationSineQua1997 trained participants with either 8, 20, or 50 unique items (all receiving the same total number of training trials). These larger sets of unique items, combined with the fact that participants achieved near ceiling level performance by the end of training - may have made it more difficult to observe any between-group differences of training variation in their study. @vandamMappingShapeVisuomotor2015 's variability manipulation was more similar to our own, as they trained participants with either 2 or 5 unique items. However, although the mapping between their input stimuli and motor responses was technically linear, the input dimension was more complex than our own, as it was defined by the degree of "spikiness" of the input shape. This entirely arbitrary mapping also would have preculded any sense of a "0" point, which may partially explain why neither of their training conditions were able to extrapolate linearly in the manner observed in the current study or in @deloshExtrapolationSineQua1997.

@@

While the present study provides valuable insights into the influence of training variability on visuomotor function learning and extrapolation, there are several limitations that should be flagged. 
First, although the constant training group never had experience from a velocity band closer to the extrapolation bands than the varied group, they always had a three times more trials with the nearest velocity band. Such a difference may be an unavoidable consequence of varied vs. constant design which match the total number of training trials between the two groups. However in order to more carefully tease apart the influence of variability from the influence of frequency/repetition effects, future research could explore alternative designs that maintain the variability manipulation while equating the amount of training on the nearest examples across conditions, such as by increasing the total number of trials for the varied group.
Another limitation is that the testing stage did not include any interpolation items, i.e. the participants tested only from the training bands they experienced during training, or from extrapolation bands. The absence of interpolation testing makes it more difficult to distinguish between the effects of training variability on extrapolation specifically, as opposed to generalization more broadly. Of course, the nature of the constant training condition makes interpolation testing impossible to implement, however future studies might compare a training regimes that each include at least 2 distinct items, but still differ in total amount of variability experienced, which would then allow groups to be compared in terms of both interpolation and extrapolation testing. 
Finally, the task employed in the present study consisted of only a linear, positive function. Previous work in human function learning has repeatedly shown that such functions are among the easiest to learn, but that humans are nonetheless capable of learning negative, non-linear, or discontinuous functions [@deloshExtrapolationSineQua1997; @mcdanielPredictingTransferPerformance2009; @kalishLearningExtrapolatingPeriodic2013; @busemeyerLearningFunctionalRelations1997]. It thus remains an open question as to whether the influence of training variability might interact with various components of the to-be-learned function.

>>>>

---
title: Impact of Training Variability on Visuomotor Function Learning and Extrapolation
author:
- name: Thomas E. Gorman
  affiliations: 
  - name:  Department of Psychological and Brain Sciences, Indiana University-Bloomington, USA
    affiliation-url: https://psych.indiana.edu/
  - name:  Program in Cognitive Science, Indiana University-Bloomington, USA
    affiliation-url: https://cogs.indiana.edu/index.html
  url: https://tegorman13.github.io/
  email: tegorman@iu.edu
  orcid: "0000-0001-5366-5442"
- name: Robert L. Goldstone
  affiliation: Department of Psychological and Brain Sciences, Indiana University-Bloomington
  affiliation-url: https://psych.indiana.edu/
  url: https://pc.cogs.indiana.edu/
  email: rgoldsto@indiana.edu
  orcid: "0000-0001-8357-8358"
# abstract: |
#   In project 1, we applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, we will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. Project 2 also places a greater emphasis on extrapolation performance following training - as varied training has often been purported to be particularly beneficial in such situations. 
keywords:
  - Learning Generalization
  - Function Learning
  - Visuomotor learning
  - Training Variability
page-layout: full
code-repo: "Access the code, data, and analysis at <https://github.com/tegorman13/htw>"
toc: false
---

@@

Our findings also diverge from the two previous studies that cleanly manipulated the variability of training items in a function learning task [@deloshExtrapolationSineQua1997; @vandamMappingShapeVisuomotor2015], although the varied training condition of @vandamMappingShapeVisuomotor2015 also exhibited less learning, neither of these previous studies observed any difference between training conditions in extrapolation to novel items. Like @deloshExtrapolationSineQua1997 , our participants exhibited above chance extrapolation/discrimination of novel items, however they observed no difference between any of their three training conditions. A noteworthy difference difference between our studies is that @deloshExtrapolationSineQua1997 trained participants with either 8, 20, or 50 unique items (all receiving the same total number of training trials). These larger sets of unique items, combined with the fact that participants achieved near ceiling level performance by the end of training, may have made it more difficult to observe any between-group differences of training variation in their study. @vandamMappingShapeVisuomotor2015 's variability manipulation was more similar to our own, as they trained participants with either 2 or 5 unique items. However, although the mapping between their input stimuli and motor responses was technically linear, the input dimension was more complex than our own, as it was defined by the degree of "spikiness" of the input shape. This entirely arbitrary mapping also would have precluded any sense sensible "0" point, which may partially explain why neither of their training conditions were able to extrapolate linearly in the manner observed in the current study or in @deloshExtrapolationSineQua1997.

@@

While the present study provides valuable insights into the influence of training variability on visuomotor function learning and extrapolation, there are several limitations that should be flagged. 
First, although the constant training group never had experience from a velocity band closer to the extrapolation bands than the varied group, they always had a three times more trials with the nearest velocity band. Such a difference may be an unavoidable consequence of a varied vs. constant design which matches the total number of training trials between the two groups. However, in order to more carefully tease apart the influence of variability from the influence of frequency/repetition effects, future research could explore alternative designs that maintain the variability manipulation while equating the amount of training on the nearest examples across conditions, such as by increasing the total number of trials for the varied group.
Another limitation is that the testing stage did not include any interpolation items, i.e. the participants were tested only from the training bands they experienced during training, or from extrapolation bands. The absence of interpolation testing makes it more difficult to distinguish between the effects of training variability on extrapolation specifically, as opposed to generalization more broadly. Of course, the nature of the constant training condition makes interpolation testing impossible to implement, however future studies might compare training regimes that each include at least 2 distinct items, but still differ in the total amount of variability experienced, which would then allow groups to be compared in terms of both interpolation and extrapolation testing. 
Finally, the task employed in the present study consisted of only a linear, positive function. Previous work in human function learning has repeatedly shown that such functions are among the easiest to learn, but that humans are nonetheless capable of learning negative, non-linear, or discontinuous functions [@deloshExtrapolationSineQua1997; @mcdanielPredictingTransferPerformance2009; @kalishLearningExtrapolatingPeriodic2013; @busemeyerLearningFunctionalRelations1997]. It thus remains an open question as to whether the influence of training variability might interact with various components of the to-be-learned function.

### 2024-07-09 Sections/Discussion.qmd
re-render with updated limitations
maybe
turn off external
final edits from Rob's doc

*Experimental Result Summary*

Across three experiments, we investigated the impact of training variability on learning and extrapolation in a visuomotor function learning task. In Experiment 1, participants in the varied training condition, who experienced a wider range of velocity bands during training, showed lower accuracy at the end of training compared to those in the constant training condition.

Crucially, during the testing phase, the varied group exhibited significantly larger deviations from the target velocity bands, particularly for the extrapolation bands that were not encountered during training. The varied group also showed less discrimination between velocity bands, as evidenced by shallower slopes when predicting response velocity from target velocity band.

@@

Our findings also diverge from the two previous studies to cleanly manipulate the variability of training items in a function learning task [@deloshExtrapolationSineQua1997; @vandamMappingShapeVisuomotor2015], although the varied training condition of @vandamMappingShapeVisuomotor2015 also exhibited less learning, neither of these previous studies observed any difference between training conditions in extrapolation to novel items. Like @deloshExtrapolationSineQua1997 , our participants exhibited above chance extrapolation/discrimination of novel items, however they observed no difference between any of their three training conditions. A noteworthy difference difference between our studies is that @deloshExtrapolationSineQua1997 trained participants with either 8, 20, or 50 unique items (all receiving the same total number of training trials). These larger sets of unique items, combined with the fact that participants achieved near ceiling level performance by the end of training - may have made it more difficult to observe any between-group differences of training variation in their study. @vandamMappingShapeVisuomotor2015 's variability manipulation was more similar to our own, as they trained participants with either 2 or 5 unique items. However, although the mapping between their input stimuli and motor responses was technically linear, the input dimension was more complex than our own, as it was defined by the degree of "spikiness" of the input shape. This entirely arbitrary mapping also would have preculded any sense of a "0" point, which may partially explain why neither of their training conditions were able to extrapolate linearly in the manner observed in the current study or in @deloshExtrapolationSineQua1997.

*Modeling Summary*
EXAM is the best model for both groups, but EXAM does relatively good at accounting for the constant group. May have seemed counterintuitive, if one assumed that multiple, varied, examples were necessary to extract a rule. But, EXAM is not a conventional rule model - it doesn't require explictly abstract of a rule, but rather the rule-based response occurs during retrieval. The constant groups formation of a single, accurate, input-output association, in combination with the usefulness of the zero point, may have been sufficient for EXAM, and the constant group,  to perform well. One concern may have been that the assumption of participants making use of the zero point turned the extrapolation problem into an interpolation problem - however this concern is ameliorated by the consistency of the results across both the original and reverse order conditions.

@@

While the present study provides valuable insights into the influence of training variability on visuomotor function learning and extrapolation, there are several limitations that should be flagged. 
First, although the constant training group never had experience from a velocity band closer to the extrapolation bands than the varied group, they always had a three times more trials with the nearest velocity band. Such a difference may be an unavoidable consequence of varied vs. constant design which match the total number of training trials between the two groups. However in order to more carefully tease apart the influence of variability from the influence of frequency/repetition effects, future research could explore alternative designs that maintain the variability manipulation while equating the amount of training on the nearest examples across conditions, such as by increasing the total number of trials for the varied group.
Another limitation is that the testing stage did not include any interpolation items, i.e. the participants tested only from the training bands they experienced during training, or from extrapolation bands. The absence of interpolation testing makes it more difficult to distinguish between the effects of training variability on extrapolation specifically, as opposed to generalization more broadly. Of course, the nature of the constant training condition makes interpolation teseting impossible to implement, however future studies might compare a training regimes that each include at least 2 distinct items, but still differ in total amount of variability experienced, which would then allow groups to be compared in terms of both interpolation and extrapolation testing. 
Finally, the task employed in the present study consisted of only a linear, positive function. Previous work in human function learning has repeatedly shown that such functions are among the easiest to learn, but that humans are nonetheless capable of learning negative, non-linear, or discontinuous functions [@deloshExtrapolationSineQua1997; @mcdanielPredictingTransferPerformance2009; @kalishLearningExtrapolatingPeriodic2013; @busemeyerLearningFunctionalRelations1997]. It thus remains an open question as to whether the influence of training variability might interact with various components of the to-be-learned function. `

>>>>

Across three experiments, we investigated the impact of training variability on learning and extrapolation in a visuomotor function learning task.

In Experiment 1, participants in the varied training condition, who experienced a wider range of velocity bands during training, showed lower accuracy at the end of training compared to those in the constant training condition. Crucially, during the testing phase, the varied group exhibited significantly larger deviations from the target velocity bands, particularly for the extrapolation bands that were not encountered during training. The varied group also showed less discrimination between velocity bands, as evidenced by shallower slopes when predicting response velocity from target velocity band.

@@

Our findings also diverge from the two previous studies that cleanly manipulated the variability of training items in a function learning task [@deloshExtrapolationSineQua1997; @vandamMappingShapeVisuomotor2015], although the varied training condition of @vandamMappingShapeVisuomotor2015 also exhibited less learning, neither of these previous studies observed any difference between training conditions in extrapolation to novel items. Like @deloshExtrapolationSineQua1997 , our participants exhibited above chance extrapolation/discrimination of novel items, however they observed no difference between any of their three training conditions. A noteworthy difference difference between our studies is that @deloshExtrapolationSineQua1997 trained participants with either 8, 20, or 50 unique items (all receiving the same total number of training trials). These larger sets of unique items, combined with the fact that participants achieved near ceiling level performance by the end of training, may have made it more difficult to observe any between-group differences of training variation in their study. @vandamMappingShapeVisuomotor2015 's variability manipulation was more similar to our own, as they trained participants with either 2 or 5 unique items. However, although the mapping between their input stimuli and motor responses was technically linear, the input dimension was more complex than our own, as it was defined by the degree of "spikiness" of the input shape. This entirely arbitrary mapping also would have precluded any sense sensible "0" point, which may partially explain why neither of their training conditions were able to extrapolate linearly in the manner observed in the current study or in @deloshExtrapolationSineQua1997.

@@

While the present study provides valuable insights into the influence of training variability on visuomotor function learning and extrapolation, there are several limitations that should be flagged. 
First, although the constant training group never had experience from a velocity band closer to the extrapolation bands than the varied group, they always had a three times more trials with the nearest velocity band. Such a difference may be an unavoidable consequence of a varied vs. constant design which matches the total number of training trials between the two groups. However, in order to more carefully tease apart the influence of variability from the influence of frequency/repetition effects, future research could explore alternative designs that maintain the variability manipulation while equating the amount of training on the nearest examples across conditions, such as by increasing the total number of trials for the varied group.
Another limitation is that the testing stage did not include any interpolation items, i.e. the participants were tested only from the training bands they experienced during training, or from extrapolation bands. The absence of interpolation testing makes it more difficult to distinguish between the effects of training variability on extrapolation specifically, as opposed to generalization more broadly. Of course, the nature of the constant training condition makes interpolation testing impossible to implement, however future studies might compare training regimes that each include at least 2 distinct items, but still differ in the total amount of variability experienced, which would then allow groups to be compared in terms of both interpolation and extrapolation testing. 
Finally, the task employed in the present study consisted of only a linear, positive function. Previous work in human function learning has repeatedly shown that such functions are among the easiest to learn, but that humans are nonetheless capable of learning negative, non-linear, or discontinuous functions [@deloshExtrapolationSineQua1997; @mcdanielPredictingTransferPerformance2009; @kalishLearningExtrapolatingPeriodic2013; @busemeyerLearningFunctionalRelations1997]. It thus remains an open question as to whether the influence of training variability might interact with various components of the to-be-learned function.
